---
title: Test Post to Explore Formatting
author: John Goldin
date: '2020-02-16'
slug: []
categories: []
tags: []
type: post
subtitle: ''
image: ''
draft: yes
---

## Test Post

If I have set this correctly (draft: yes), this post should only appear
in my local RStudio session and not get into the distribution from Netlify.

If anyone except me sees this, then I have made a mistake!


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE, out.width = "100%",
                      knitr.table.format = "html")
# knitr.table.format is for kable. Needed this to get blogdown table to format correctly.
```
```{r libraries, message = FALSE, warning = FALSE}
library(tidyverse, quietly = TRUE)
library(knitr, quietly = TRUE)
library(scales, quietly = TRUE)
library(kableExtra, quietly = TRUE)
```
```{r results="asis", echo = FALSE}
cat("
<style>
code {
   color: #blue;
}
</style>
")
```
```{css echo=FALSE}
code {
  color: #97736C;
  background-color: #FAF9F9;
  font-weight: 400
}
```
Use https://htmlcolorcodes.com to find color.

Can I highlight a single word with back-quote and not get the red highlight I normally see with
code?

This is a sentence with a `filename` that is highlighted.

This is inline code: `r text_spec(colnames(mtcars)[1], color = "red")`.

Version 10

previous css block:
  font-weight: 600


```{r include_stuff FALSE}
path_saved_export <- "~/Dropbox/Programming/R_Stuff/john_vitals/Apple-Health-Data/"
path_to_healthexport1 <- "~/Documents/R_local_repos/applehealth1/R/"
# source(paste0(path_to_healthexport1, "find_timezone.R"))
# source(paste0(path_to_healthexport1, "tripit_functions.R"))
load(paste0(path_saved_export, "small_health_df.RData"))
health_df <- small_health_df
```
### The Types of Data in the Export

The counts by "type" describe the breadth and quantity of data can be viewed in Table \@ref(tab:freq-type):

```{r freq-type, echo = FALSE}
# see note on why chunk label needs hyphen: https://github.com/rstudio/distill/issues/46#issuecomment-442658763
health_df2 <- health_df %>% 
  mutate(source = case_when(
    str_detect(sourceName, "Phone") ~ "Phone",
    str_detect(sourceName, "Watch") ~ "Watch",
    str_detect(sourceName, "Lose It") ~ "Lose It!",
    TRUE ~ "Other"),
    source = factor(source, levels = c("Watch", "Phone", "Lose It!", "Other")))
  # can use clipr::write_clip(allow_non_interactive = TRUE) if want to paste table
# padding tip from https://stackoverflow.com/questions/39252915/r-knitr-kable-padding-not-working-with-format-html
health_df2 %>% 
  janitor::tabyl(type, source) %>% 
  janitor::adorn_totals("col") %>%  arrange(desc(Total)) %>% 
  janitor::adorn_totals("row") %>% # do column total after arrange
  kable(format.args = list(decimal.mark = " ", big.mark = ","),
        # table.attr='class="myTable"',
        # caption = "Frequency by Type and Data Source", format = "markdown") %>% 
        caption = "Frequency by Type and Data Source")
        
```

```

