---
title: Tracking Covid-19 in Connecticut
author: John Goldin
date: '2020-06-22'
categories:
  - COVID19
tags:
  - COVID19
slug: covid-19-in-Connecticut
type: post
draft: no
output:
#  blogdown::html_page:
  blogdown::html_page:
    toc: yes
    number_sections: false
    # figure parameters based on recommendation in Hadley's book
    fig_width: 7 
    out.width: "100%" 
    fig.align: "center"
    fig.asp: 0.618
---

```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE, #out.width = "100%",
                      knitr.table.format = "html")
# knitr.table.format is for kable. Needed this to get blogdown table to format correctly.
```
```{r libraries, echo = FALSE, message = FALSE, warning = FALSE}
# subtitle: "Updated `r format(Sys.time(), '%a %d %b %Y')`"
library(tidyverse, quietly = TRUE)
library(knitr, quietly = TRUE)
library(scales, quietly = TRUE)
library(kableExtra, quietly = TRUE)
library(sparkline)
library(geofacet) # https://hafen.github.io/geofacet/
library(sf, quietly = TRUE)
options(tigris_use_cache = TRUE)
library(tigris, quietly = TRUE)
library(tidycensus, quietly = TRUE)
library(viridis, quietly = TRUE)
library(ggrepel)
options(tigris_class = "sf")
library(lubridate, quietly = TRUE)
library(janitor)
library(fuzzyjoin) # for interval_left_join
library(broom)
library(RSocrata)
library(RcppRoll)
library(gt)
# library(formattable)
```
```{r `load_basic_data`, echo = FALSE, message = FALSE}
path_to_post <- "~/Dropbox/Programming/R_Stuff/can_i_blog_too/content/post/2020-03-29-covid19-cases-in-connecticut/"
path_to_static_for_this_post <- "~/Dropbox/Programming/R_Stuff/can_i_blog_too/static/post/2020-06-19-tracking-covid-19-in-connecticut"
path_to_ctcorona <- "~/Documents/R_local_repos/ctcorona/data/"
path_to_ctcorona_r <- "~/Documents/R_local_repos/ctcorona/R/"
load(paste0(path_to_ctcorona, "dph_datasets.RData"))
load(paste0(path_to_ctcorona, "census_population.RData"))
source(paste0(path_to_ctcorona_r, "combine_dph_towns.R"))

# from https://www.ctdatahaven.org/sites/ctdatahaven/files/UConnCPR%20Changing%20Demographics-5%20CTs%202004.pdf
# The Changing Demographics of Connecticut — 1990 to 2000. by Center for Population Research  May. 31, 2004
Five_Connecticuts <- read_delim(paste0(path_to_ctcorona, "Five_Connecticuts.txt"),"\t", escape_double = FALSE, trim_ws = TRUE) %>%
  mutate(category = factor(category, levels = c("Urban Core", "Urban Periphery",
                           "Wealthy", "Suburban", "Rural")))
# note that prisons have dominant effect on Somers and Montville
load(paste0(path_to_ctcorona, "dph_datasets.RData"))
# load effective R from https://rt.live/
rt_live <- read_csv("https://d14wlfuexuxgcm.cloudfront.net/covid/rt.csv") %>% 
  mutate(state = case_when(
    region == "CT" ~ "Connecticut",
    region == "AZ" ~ "Arizona"
  )) %>% 
  group_by(region) %>% arrange(date) %>% 
  mutate(above_one = ifelse(last(mean) > 1, "> 1", "<= 1") %>% factor(levels = c("> 1", "<= 1")),
         recent_rt = last(mean)) %>% 
  ungroup() 
```


## Latest Connecticut Data from the Department of Public Health

```{r box-score, echo = FALSE} 
#<!--more-->
recent_date <- max(dph_total$date, na.rm = TRUE)
week_ago <- recent_date - 7
if (!(week_ago %in% dph_total$date)) week_ago <- NA
yesterday <- recent_date - 1
# if (!(yesterday %in% dph_total$date)) yesterday <- NA
# if (is.na(yesterday) || (length(dph_total$deaths[dph_total$date == yesterday]))) {
#   yesterday_deaths <- ""
# } else {yesterday_deaths <- dph_total$deaths[dph_total$date == recent_date] - dph_total$deaths[dph_total$date == yesterday]}
today_cases <- dph_total %>% filter(date == recent_date) %>% 
  select(cases) %>% pluck(1, .default = NA)
yesterday_cases <- today_cases - dph_total %>% filter(date == yesterday) %>% 
  select(cases) %>% pluck(1, .default = NA)
week_ago_cases <-  today_cases - dph_total %>% filter(date == week_ago) %>% 
  select(cases) %>% pluck(1, .default = NA)
today_deaths <- dph_total %>% filter(date == recent_date) %>% 
  select(deaths) %>% pluck(1, .default = NA)
yesterday_deaths <- today_deaths - dph_total %>% filter(date == yesterday) %>% 
  select(deaths) %>% pluck(1, .default = NA)
week_ago_deaths <-  today_deaths - dph_total %>% filter(date == week_ago) %>% 
  select(deaths) %>% pluck(1, .default = NA)
today_hospital <- dph_total %>% filter(date == recent_date) %>% 
  select(hospital) %>% pluck(1, .default = NA)
yesterday_hospital <- today_hospital - dph_total %>% filter(date == yesterday) %>% 
  select(hospital) %>% pluck(1, .default = NA)
week_ago_hospital <-  today_hospital - dph_total %>% filter(date == week_ago) %>% 
  select(hospital) %>% pluck(1, .default = NA)
today_tests <- dph_total %>% filter(date == recent_date) %>% select(covid_19_pcr_tests_reported) %>% pluck(1, .default = NA)
yesterday_tests <- today_tests - dph_total %>% filter(date == yesterday) %>% select(covid_19_pcr_tests_reported) %>% pluck(1, .default = NA)
week_ago_tests <-  today_tests - dph_total %>% filter(date == week_ago) %>% select(covid_19_pcr_tests_reported) %>% pluck(1, .default = NA)
test_rate_day <- ((dph_total %>% filter(date == recent_date) %>% 
                 select(confirmedcases) %>% pluck(1, .default = NA)) -
                (dph_total %>% filter(date == yesterday) %>% 
                   select(confirmedcases) %>% pluck(1, .default = NA))) / 
  ((dph_total %>% filter(date == recent_date) %>% 
      select(covid_19_pcr_tests_reported) %>% pluck(1, .default = NA)) -
     (dph_total %>% filter(date == yesterday) %>% 
        select(covid_19_pcr_tests_reported) %>% pluck(1, .default = NA)))
test_rate_week <- ((dph_total %>% filter(date == recent_date) %>% 
                 select(confirmedcases) %>% pluck(1, .default = NA)) -
                (dph_total %>% filter(date == week_ago) %>% 
                   select(confirmedcases) %>% pluck(1, .default = NA))) / 
  ((dph_total %>% filter(date == recent_date) %>% 
      select(covid_19_pcr_tests_reported) %>% pluck(1, .default = NA)) -
     (dph_total %>% filter(date == week_ago) %>% 
        select(covid_19_pcr_tests_reported) %>% pluck(1, .default = NA)))
# next two versions don't deal with confirmed cases only
test_rate_day2 <- yesterday_cases / yesterday_tests
test_rate_week2 <- week_ago_cases / week_ago_tests
recent_nursing_date <- max(dph_nursing_cases$date, na.rm = TRUE)
nursing_previous_date <- dph_nursing_cases %>% select(date) %>% arrange(desc(date)) %>% unique()
nursing_previous_date <- nursing_previous_date$date[2]
nursing_hm_cases <- dph_nursing_cases %>% filter(date == recent_nursing_date) %>% 
  select(nh_cases) %>% pluck(1, .default = NA) %>% sum(na.rm = TRUE)
nh_previous_cases <-  nursing_hm_cases - dph_nursing_cases %>% filter(date == nursing_previous_date) %>% 
  select(nh_cases) %>% pluck(1, .default = NA) %>% sum(na.rm = TRUE)
nursing_hm_deaths <- dph_nursing_cases %>% filter(date == recent_nursing_date) %>% 
  select(nh_deaths) %>% pluck(1, .default = NA) %>% sum(na.rm = TRUE)
nh_previous_deaths <-  nursing_hm_deaths - dph_nursing_cases %>% filter(date == nursing_previous_date) %>% 
  select(nh_deaths) %>% pluck(1, .default = NA) %>% sum(na.rm = TRUE)
# use: sprintf("%+3d %%", x)
box_fmt <- function(x) {
  case_when(
    (is.na(x) | is.null(x) | (length(x) == 0)) ~ " ",
    x < 0 ~ comma(x),
    x > 0 ~ comma(x, prefix = "+"),
    TRUE ~  "0"
  )
}

###########################################################################
# Box score comes here:
box_score <- tribble(
  ~`Connecticut Covid-19`, ~`today`, ~`yesterday`, ~`week ago`,
  "Cases", today_cases, yesterday_cases, week_ago_cases,
  "Deaths", today_deaths, yesterday_deaths, week_ago_deaths,
  "Currently in Hospital", today_hospital, yesterday_hospital, week_ago_hospital,
  "Tests",    today_tests, yesterday_tests, week_ago_tests,
  " ", NA , NA, NA,
  "Nursing home cases", nursing_hm_cases, NA, nh_previous_cases,
  "Nursing home deaths", nursing_hm_deaths, NA, nh_previous_deaths
) %>% 
   mutate(yesterday = map_chr(yesterday, box_fmt),
         `week ago` = map_chr(`week ago`, box_fmt), today = ifelse(is.na(today), "", comma(today)))
```
```{r format-box-score-table, echo = FALSE}
# columns for box scores:
# town, category, population, cases, cases per 100K, recent cases per 100K, percent positive tests
# deaths, percent of deaths in nh, 
box_score %>% 
  tibble::add_row(`Connecticut Covid-19` = "Rate of positive tests", 
                  today = "", 
                  yesterday = paste0(round(test_rate_day * 100, 1.1), "%"), `week ago` = paste0(round(test_rate_week * 100, 1.1), "%"), .after = 4) %>% 
  gt() %>% 
  tab_header(title = "Latest Data on Covid-19 in Connecticut",
             subtitle = md(paste0('Should correspond to [', dph_reports$name[1],
                                  '](', dph_reports$url[1], ') from the Department of Public Health'))) %>% 
  cols_label(
    `Connecticut Covid-19` = html(""),
    today = html("total<br>to date"),
    yesterday = html("since<br>yesterday"),
    `week ago` = html("since a<br>week ago")
  ) %>% 
  cols_align(align = "right") %>% 
  cols_align(align = "left", columns = c(1)) %>% 
  tab_source_note(source_note =  md("Source: Multiple tables from [Connecticut Open Data](https://data.ct.gov/stories/s/COVID-19-data/wa3g-tfvc/)")) %>% 
    tab_footnote(
    footnote = paste0('Nursing home data are reported only weekly so the dates for "today" and a "week ago" are ', format(recent_nursing_date,"%m/%d/%Y"), " and ", format(nursing_previous_date,"%m/%d/%Y"), "."),
    locations = cells_body(
      columns = vars(`Connecticut Covid-19`),
      rows = 7)
  ) %>% 
  # kable_styling(bootstrap_options = ".table", position = "left")
  opt_row_striping(row_striping = FALSE)

# issue about css and gt table
# https://github.com/rstudio/gt/issues/547

  # tab_options(row.striping.include_table_body = FALSE)

```

 
[My post on Covid-19 back in March](https://www.johngoldin.com/2020/03/covid-19-cases-in-connecticut/) 
tracked the growth of the epidemic as it hit Connecticut.
Back at the end of March I did not realize that 
the greater New York area (including Connecticut)
would be the center of the first wave in the US. At that time what I was reading
focused on the nature of exponential growth of an epidemic. We were urged to
"flatten the curve" and that shaped the way I presented the
data in that post. I have been updating the figures
in that post each evening. During that time the [Connecticut data portal](https://data.ct.gov/stories/s/COVID-19-data/wa3g-tfvc/)
has expanded considerably, and I've built [some R code](https://github.com/johngoldin/ctcorona/blob/master/R/daily_ct_stats.R) 
that let's me easily update
that post. I'll use the same data infrastructure for this post as well.
This post is more oriented toward tracking the current status of 
the epidemic in Connecticut.

```{r 'new-cases-plot', echo = FALSE, eval = TRUE}
## This imaged will be saved in the /post/*_files/ directory
## Use echo = FALSE if you want to hide the code for making the plot
exec_orders <- tibble(
  date = as.Date(c("2020-03-10", "2020-03-17", "2020-03-23", "2020-03-26", "2020-04-11", "2020-04-20", "2020-05-20", "2020-06-01", "2020-06-17")),
  label = c("prohibit large\ngatherings", "cancel\nclasses", "restrict\nbusiness", "5 or\nfewer",
            "covid-19\ndeath\nchanged", "masks\nrequired",
            "phase 1\nreopen", "hair\nreopen", "phase 2\nreopen"),
  fudge = c(1, 0.4, 0.7, 0.9, 0.8, 0.9, 0.8, 0.8, 1)
) %>% 
  left_join(dph_total %>% select(date, rnew_cases, rnew_deaths, cases, deaths))

# try a version with understated bars
p_new_cases <- ggplot(data = dph_total %>% 
              # filter(county == "Fairfield") %>% 
              mutate(wday  = factor(weekdays(date),
                                  levels = c("Monday", "Tuesday",
                                             "Wednesday", "Thursday", "Friday",
                                             "Saturday","Sunday"))), 
            aes(x = date, y = new_cases)) +
  geom_col(fill = "gray95", colour = "gray90") + 
  # scale_fill_viridis(discrete = TRUE, begin = 0.5, end = 0.7, option = "viridis") + 
  scale_y_continuous(labels = comma, limits = c(0, NA)) +
  scale_x_date(date_breaks = "2 weeks", date_minor_breaks = "1 week", date_labels = "%b %d") +
  geom_line(aes(y = rnew_cases), size = 1) +
  theme_minimal() + theme(legend.position = "none") +
  xlab(NULL) + ylab("New Cases Per Day") +
  geom_text(data = exec_orders %>% filter(date <= max(dph_total$date, na.rm = TRUE)), 
            aes(y = (rnew_cases + 500) * fudge , label = label),
            hjust = 0.5, vjust = 0, size = 3, colour = "black", nudge_y = 0) +
  ggtitle("Connecticut New Covid-19 Cases (with 7-day rolling average)")

p_daily_deaths <- ggplot(data = dph_total %>% 
              # filter(county == "Fairfield") %>% 
              mutate(wday  = factor(weekdays(date),
                                  levels = c("Monday", "Tuesday",
                                             "Wednesday", "Thursday", "Friday",
                                             "Saturday","Sunday"))), 
            aes(x = date, y = new_deaths)) +
  geom_col(fill = "gray95", colour = "gray90") + 
  # scale_fill_viridis(discrete = TRUE, begin = 0.5, end = 0.7, option = "viridis") + 
  scale_y_continuous(limits = c(0, NA)) +
  scale_x_date(date_breaks = "2 weeks", date_minor_breaks = "1 week", date_labels = "%b %d") +
  geom_line(aes(y = rnew_deaths), size = 1) +
  theme_minimal() + theme(legend.position = "none") +
  xlab(NULL) + ylab("Deaths Per Day") +
  geom_text(data = exec_orders %>% filter(date <= max(dph_total$date, na.rm = TRUE)), 
            aes(y = (rnew_deaths + 75) * fudge , label = label),
            hjust = 0.5, vjust = 0, size = 3, colour = "black", nudge_y = 0) +
  ggtitle("Connecticut Covid-19 Daily Deaths (with 7-day rolling average)")

r_data_ct <- rt_live %>% filter(region == "CT")
r_data_az <- rt_live %>% filter(region == "AZ")
# colors from colorbrewer:  https://colorbrewer2.org/#type=diverging&scheme=PRGn&n=3
p_rt_live <- ggplot(data = r_data_ct, aes(x = date, y = mean)) +
  geom_ribbon(aes(ymin = lower_80, ymax = upper_80), fill = "grey83") +
  geom_ribbon(data = r_data_az, aes(ymin = lower_80, ymax = upper_80), fill = "grey83") +
  geom_line(aes(colour = above_one), size = 1) + theme_minimal() + 
  geom_hline(yintercept = 1, colour = "gray") +
  geom_line(data = r_data_az, aes(colour = above_one), size = 1) + theme_minimal() + 
  geom_text(data = rt_live %>% filter(region %in% c("CT", "AZ")) %>%
              group_by(region, state) %>% arrange(region, state, date) %>%
              summarise(date = min(date), recent_rt = last(recent_rt)),
            aes(label = state, x = ymd("2020-03-23"), y = recent_rt), hjust = 1, size = 3, nudge_x = 2) +
  theme(legend.position = "none") +
  theme(axis.title.y = element_text(angle=0, vjust = 0.5)) +
  scale_color_manual(values = c("<= 1" = "#7fbf7b", "> 1" = "#af8dc3")) +
  scale_y_continuous(limits = c(0.5, 2.5), breaks = c(0.5, 1, 1.5, 2.0)) +
                     #minor_breaks = seq(0.2, 1.4, 0.2)) +
  scale_x_date(date_breaks = "2 weeks", date_minor_breaks = "1 week", 
               date_labels = "%b %d", 
               limits = c(min(dph_total$date), max(dph_total$date))) +
  xlab(NULL) + ylab(expression("R"[t])) +
  # see: http://strata.uga.edu/8370/rtips/subscriptsSuperscripts.html
  labs(title = expression(paste("Trend in ", R[t], " (as computed by https://rt.live/)")))
```
```{r hospitalizations, echo = FALSE, out.width = "70%", fig.align = "center"}
p_hospitalizations <- ggplot(data = dph_total %>% filter(!is.na(hospital)),
                    aes(x = date, y = hospital)) +
  geom_line() + xlab(NULL) + # geom_point() + 
  scale_x_date(date_breaks = "2 weeks", date_minor_breaks = "1 week", 
               date_labels = "%b %d", 
               limits = c(min(dph_total$date), NA)) +
  theme_minimal() + scale_y_continuous(labels = comma) +
  theme(legend.position = "none") +
  ylab("in hospital") +
  labs(
    title = "COVID19 Patients Currently in Hospital in Connecticut",
    subtitle = NULL,
    caption = "Source: Connecticut Hospital Association, via CT DPH"
  )
```
```{r geofacets, echo = FALSE}
#  <span style="color: #ff0000">January 30, 2011</span> > 1 #af8dc3 <= 1 #7fbf7b
p_rt_map <- ggplot(data = rt_live %>% select(-state) %>% 
         rename(state = region), aes(x = date, y = mean)) + 
  xlab(NULL) + ylab(NULL) + 
  theme_bw() + 
  geom_ribbon(aes(ymin = lower_80, ymax = upper_80), fill = "grey80") +  # higher grey number is lighter
  geom_hline(yintercept = 1, colour = "grey50") +
  geom_line(aes(colour = above_one), size = 1) + theme(legend.position = "left") +
  scale_color_manual(values = c("<= 1" = "#7fbf7b", "> 1" = "#af8dc3")) +
  scale_y_continuous(limits = c(0.5, 1.8), breaks = c(0.5, 1, 1.5)) +
                     #minor_breaks = seq(0.2, 1.4, 0.2)) +
  scale_x_date(date_breaks = "2 months", date_minor_breaks = "1 month", 
               date_labels = "%b") + 
  theme(axis.title.y = element_text(angle=0, vjust = 0.5, size = 6)) +
  theme(axis.text.x = element_text(size=5)) +
  # labs(title = paste0("Trends in ", expression("R"[t]), "(as computed by https://rt.live/)")) +
  labs(title = expression(paste("Trend in ", R[t], " by State (as computed by https://rt.live/)")),
       colour = expression(paste("Recent ", R[t]))) +
  facet_geo(~ state)

```
## Tracking the Virus in Connecticut

This post focuses on what is happening currently. The aim is to
use the recent trend to get some idea of the near future. The first
plot will show the average number of new cases reported each day along
with a line that displays the rolling seven-day average. Typically there 
are day-of-the-week effects in the reporting so it's best to focus
on the seven-day average.

### New Covid-19 Cases in Connecticut

The plot shows the history of new cases and also shows the 
actions required by some of Governor Lamont's executive orders. 
<br/><br/><br/>
```{r p-new-cases, echo = FALSE, fig.cap = "New Cases"}
print(p_new_cases)

```
<br/><br/><br/>
```{r p-rt-live, echo = FALSE, fig.asp = 0.3, fig.cap = "Estimated R<sub>t</sub>"}
print(p_rt_live)
```
<br/>

### Covid-19 Deaths in Connecticut

```{r p-daily-deaths, echo = FALSE, fig.cap = "Daily Deaths"}
print(p_daily_deaths)

```
<br/>

### Covid-19 Hospitalizations in Connecticut

```{r p-hospitalizations, echo = FALSE, fig.cap = "Hospitalizations for Covid-19"}
print(p_hospitalizations)
```
<br/><br/>
The number of new cases peaked in mid April and then began to
decrease rapidly, although not as rapidly as it increased. 
A key issue is how fast epidemic is expanding or contracting.
One indicator of that is to estimate the
parameter R<sub>t</sub>, the average number of people 
who become infected by an infectious person. 
When R<sub>t</sub> is greater than 1 the number of daily new cases is
increasing, If R<sub>t</sub> is less than 1, it is decreasing.
See [this site](https://rt.live/) for
an estimate by state of the effective value for 
R<sub>t</sub>. The site
is described in [this article](https://www.vox.com/recode/2020/4/21/21227855/coronavirus-spreading-by-state-instagram-effective-reproduction-rate) at Vox.com.
I don't claim to have the expertise to evalute the
quality of the calculation of R at the
[rt.live](https://rt.live/) site, but it seems to jive
with other information about the pattern of the epidemic
among states.

As of June (and a month
previous), Connecticut, New York, and New Jersey have among the lowest
values for R<sub>t</sub> in the country. This is good news for us
in Connecticut. I have included R<sub>t</sub> for Arizona as
well as Connecticut to provide a contrast with
a state that has lately shown [signs of a growing outbreak](https://www.fox10phoenix.com/news/arizona-medical-experts-alarmed-over-surge-in-covid-19-cases).
The goal in Connecticut and in the entire New York City area is to
keep R<sub>t</sub> well under 1. As of `r format(max(rt_live$date),"%m/%d/%Y")` R<sub>t</sub> is
`r round(rt_live$mean[(rt_live$date == max(rt_live$date)) & (rt_live$region == "CT")], 2)`. On the other hand, in Arizona R<sub>t</sub> is about `r round(rt_live$mean[(rt_live$date == max(rt_live$date)) & (rt_live$region == "AZ")], 2)`.

## Estimate of R<sub>t</sub> by State (from [rt.live](https://rt.live/))

The next display shows the trend of R<sub>t</sub> in each state
arranged according to geographic position in the US. Remember
that R<sub>t</sub> is a measure of the rate of increase. When R<sub>t</sub> is greater than 1
the number of daily new cases is
increasing, If R<sub>t</sub> is less than 1, it is decreasing. 
The color of the R line shows whether the most recent estimate of R for
a state is <span style="color: #af8dc3">above 1 (purple line)</span>
or <span style="color: #7fbf7b">equal or below 1 (green line)</span>.

R is estimated from the state data and the gray area around the line
indicates the statistical uncertainty of the estimat showing
an 80% confidence interval. Note that this display shows only whether
transmission seems to be growing or shrinking. Not shown is whether the
volume of current cases are at a low or high level.

<br/><br/><br/>
```{r p-rt-map, echo = FALSE, fig.cap = "Estimate of R<sub>t</sub> by State", fig.asp = 0.8, out.width = "90%"}
#  <span style="color: ##af8dc">above 1 (purple line)</span> > 1 #af8dc <= 1 #7fbf7b
print(p_rt_map)
```
<br/><br/><br/>

## Deaths and Cases by Age

<div style="float: left; width: 50%; padding: 5PX;">
```{r deaths-by-age2, echo = FALSE}

#```{r deaths-by-age2, echo = FALSE, out.width = '50%', fig.show = 'hold', fig.cap = "Relationship of Age to Covid-19 Deaths andCases", fig.align = "default"}
ggplot(data = dph_age %>% filter(date == max(date)), 
       aes(x = agegroups, y = deaths)) + 
  geom_col() + 
  labs(title = "Covid-19 Deaths in Connecticut by Age") +
  xlab("Age Groups") + ylab(NULL) +
  scale_y_continuous(labels = comma) +
  theme_minimal() + theme(panel.grid.major.x = element_blank())
```
</div>
<div style="float: right; width: 50%; padding: 5PX;">
```{r deaths-per-100k2, echo = FALSE}
ggplot(data = dph_age %>% filter(date == max(date)),
       aes(x = agegroups, y = as.numeric(per_100k))) + 
  geom_col() + 
  labs(title = "Covid-19 Cases by Age in Connecticut per 100K Population") +
  xlab("Age Groups") + ylab("Cases per 100K of Population") +
  scale_y_continuous(labels = comma) +
  theme_minimal() + theme(panel.grid.major.x = element_blank())
```
</div>
<br/>

Deaths increase with age, as has been well known since the beginning of the epidemic. For cases, the number of cases per population stays relatively
flat from 20 to 79 and rises very rapidly for the oldest category. I haven't
examined this data closely, but I assume this pattern is related to the
high rates of mortality in nursig homes. For the age range from 60 to 79 there
probably are a lot of seniors who are going to a great deal of trouble
avoiding places and situations where there is risk of catching the virus.
But individuals in nursing homes don't have that much control over their
own lives. I would guess that the older population outside of nursing
homes has a lower rate of catching the virus and the populaton inside
nursing homes has a higher rate. The two trends may average out to a rate that
is the same as middle aged adults. This is just speculation. I haven't dug
into the data to say anthing firm about this

## Nursing Homes

```{r nursing-homes, echo = FALSE}
nursing_home_totals <- dph_nursing_cases %>% filter(date == max(date)) %>% 
  summarise(cases = sum(nh_cases, na.rm = TRUE), deaths = sum(nh_deaths, na.rm = TRUE))
```
Nursing homes have been hard hit by the epidemic. It's hard to control  coronavirus outbreak
in a group living situation, which includes prisons and cruises ship. On top of that,
Covid-19 is especially serious for the elderly. The combination in nursing homes
has been deadly. There have been `r comma(nursing_home_totals$deaths)` deaths in Connecticut 
Nursing Homes which makes `r paste0(round((nursing_home_totals$deaths/dph_total$deaths[dph_total$date == max(dph_total$date)]) * 100, 1), "%")` of the total of `r comma(dph_total$deaths[dph_total$date == max(dph_total$date)])` deaths from Covid-19.

```{r prisons, echo = FALSE}
# see wikipedia entry: https://en.wikipedia.org/wiki/List_of_Connecticut_state_prisons
# Locations in Connecticut
# Bridgeport Correctional Center (inmate population 712)
# Brooklyn Correctional Institution (inmate population 454) Brooklyn
# Cheshire Correctional Institution (inmate population 1322)
# Corrigan-Radgowski Correctional Center (inmate population 760) Montville
# Garner Correctional Institution (inmate population 584) Newtown
# Hartford Correctional Center (inmate population 933)
# MacDougall-Walker Correctional Institution (inmate population 1522) Suffield
# Manson Youth Institution (inmate population 322) Cheshire
# New Haven Correctional Center (inmate population 718)
# Northern Correctional Institution (inmate population 135) Somers
# Osborn Correctional Institution (inmate population 1322)  Somers
# Robinson Correctional Institution (inmate population 1449) Enfield
# Willard-Cybulski Correctional Institution (inmate population 1142) Enfield
# York Correctional Institution (inmate population 898) Niantic (women)
```

## Prisons

Prisons even more than nursing homes create a situation
where it is hard to limit exposure among individuals living together. We
have less information about prisons than about nursing homes. The
Department of Corrections (DOC) posts some statistics on 
[total Covid19 rates among prisoners and staff](https://data.ct.gov/Public-Safety/COVID-19-in-Correctional-Facilities/6t8i-du3u) 
via the state open data site and some limited information of 
[rates by individual facility](https://portal.ct.gov/DOC/Common-Elements/Common-Elements/Health-Information-and-Advisories) on their web site,
but the data are not as detailed or as accessible as the data on nursing
homes.

As of `r format(max(doc_covid$date, na.rm = TRUE), "%m/%d/%Y")`, 
`r scales::comma(max(doc_covid$inmates_positive, na.rm = TRUE))` inmates
have tested positive for Covid-19. At the peak of the epidemic, 
`r scales::comma(max(doc_covid$northern_med, na.rm = TRUE))` inmates were
isolated at the Nothern Correctional Institution medical facilty.
The most recent report is that 
`r scales::comma(doc_covid$inmate_deaths[doc_covid$date == max(doc_covid$date)])`
died, and currently 
`r scales::comma(doc_covid$northern_med[doc_covid$date == max(doc_covid$date)])`
Covid-19 inmates remain in that medical facility. 

Unlike nursing homes, the number of deaths from 
Covid-19 in prisons is relatively low. Presumably that
is because the prison population has many fewer 
elderly people who are expecially vulnerable to 
the effects of Covid-19.

In addition to prisoners, the Department of Corrections reports that 
`r scales::comma(max(doc_covid$staff_positive, na.rm = TRUE))`
staff have tested positive for Covid-19, of whom
`r scales::comma(max(doc_covid$staff_returned, na.rm = TRUE))`
have been cleared to return to work.

One cannot easily relate the number of cases reported by the 
Department of Corrections to the town by town counts reported
by the Department of Public Health. It's clear that for small
rural towns the prison cases may swamp the community counts in those
towns. For example, in the town of Brooklyn in Windham County
(population about 8,200)
the DPH town stats between June 17 and June 18
showed the number of cases
jumped from 24 to 116.
That's the home of the Brooklyn Correctional Institutin
and I have to believe that site dominated those numbers
and that the process by which the Department of Corrections
repoted test results to the Department of Public Health
affected the magnitude of the one-day jump.

In effect there is a separate epidemic occurring inside
the prison system and the pace of that epidemic does not
necessarily correspond to trends in the wider community.
For the purposes of tracking Covid-19 in the population
of Connecticut generally, in some cases I will exclude some
of the small rural towns with prisons such as Somers,
Montville, and Brooklyn. But it's not practical for
me to segregate the effect of prisons on counts for
larger towns such as New Haven, Cheshire, Hartford,
and others. As of 6/16/2020, the DOC reported a cumulative total
of 117 cases amongs prisoners in Bridgeport,
125 in New Haven, 94 in Hartford, and 21 in Cheshire.
Enfield is a tricky case. There are two DOC facilities
in Enfield and as of 6/16/2020 DOC reported 265
cases there. Enfield has a population of about 44,000,
but the prison cases may have a significant effect
on the picture of Covid-19 in Enfield. Similar to Brooklyn,
one can see likely effects in the reports. On
6/10/20 DPH reported a total of 447 Covid-19 cases
in Enfield and the next day on 6/11/2020 they
reported 546, and increase of 22% in one day. 
That's highly likely to be an effect of
how cases were reported from the DOC facilities in Enfield
and how those reports reached the Department of Public Health.


## Variations by County in Connecticut

```{r county-trends, echo = FALSE}
county_centroid <- st_centroid(county_geometries) %>%
  left_join(dph_counties %>% filter(date == max(date, na.rm = TRUE)) %>%
              select(county, rnew_cases_per1k, cases, deaths, rnew_deaths_per1k), by = "county")
county_max <- dph_counties %>% group_by(county) %>% summarise(max = max(rnew_cases_per1k, na.rm = TRUE)) %>% arrange(desc(max))
for_plot1 <- dph_counties %>% 
  select(date, county, `new cases` = rnew_cases_per1k, deaths = rnew_deaths_per1k) %>% 
  mutate(county = factor(county, levels = county_max$county)) %>% 
  pivot_longer(cols = c(`new cases`, deaths)) %>% 
  mutate(the_measure = factor(name, levels = c("new cases", "deaths")))
p_county_per100k <- ggplot(data = for_plot1, aes(x = date, y = value * 100)) +
  geom_line(aes(colour = county)) +
  scale_x_date(date_breaks = "1 month", date_minor_breaks = "1 week", 
               date_labels = "%b %d") + 
  theme_minimal() + 
  xlab(NULL) + ylab("Per 100K Population") +
  facet_wrap(~ the_measure)
  # facet_wrap(~ name, scales = "free_y")
for_plot2 <- dph_counties %>% 
  select(date, county, cases = rcases, deaths = rdeaths) %>% 
  mutate(county = factor(county, levels = county_max$county)) %>% 
  pivot_longer(cols = c(cases, deaths)) %>% 
  mutate(the_measure = factor(name, levels = c("new cases", "deaths")))
# p_county_cum <- ggplot(data = for_plot2, aes(x = date, y = value)) +
#   geom_line(aes(colour = county)) +
#   scale_x_date(date_breaks = "1 month", date_minor_breaks = "1 week", 
#                date_labels = "%b %d") + 
#   theme_minimal() + 
#   xlab(NULL) + ylab("Cumulative Count (7-day rolling average)") +
#   # facet_wrap(~ the_measure)
#   facet_wrap(~ name, scales = "free_y")

modified_counties <- dph_towns %>% 
  filter(date == max(dph_towns$date, na.rm = TRUE),
         !town %in% c("Montville", "Somers", "Brooklyn")) %>% # and maybe Enfield.  Prison towns
  group_by(county) %>% 
  summarise(across(c(total_pop, rnew_cases, current_cases, rnew_deaths, cases, deaths), sum, na.rm = TRUE)) %>% 
  mutate(across(c(rnew_cases, current_cases, cases), ~ .x / (total_pop / 100000), .names = "{col}_per100k"))
p_county_map_current_cases <- ggplot(data = county_geometries %>%
            left_join(modified_counties %>% 
                        select(county, `current per 100k` = current_cases_per100k),
                      by = "county"),
             aes(geometry = geometry)) +
  geom_sf(# need aes(geometry = geometry) because joined sf to a tibble and so lost sf
          aes(fill = `current per 100k`, geometry = geometry)) +
  scale_fill_gradient(low = "white", high = "grey") +
  geom_sf_text(aes(label = paste0(county, "\n", round(`current per 100k`, 1), " per 100K")), color = "black", size = 3) +
  coord_sf(datum = NA, label_axes = "----") +
  xlab("") + ylab("") + theme_minimal()  +
  # theme(legend.position = "none") +
  labs(title = "Current Cases (total of last 14 days) by Connecticut County",
       subtitle = "shading shows current cases per 100K of population as of most recent report\n(without Somers, Brooklyn, and Montville)",
       caption = "Sources: CT DPH, US Census, tidycensus package")
p_county_map_cum_cases <- ggplot(data = county_geometries %>%
            left_join(modified_counties %>% 
                        select(county, `cases per 100k` = cases_per100k),
                      by = "county"),
             aes(geometry = geometry)) +
  geom_sf(aes(fill = `cases per 100k`, geometry = geometry)) +
  scale_fill_gradient(low = "white", high = "grey") +
  geom_sf_text(aes(label = paste0(county, "\n", scales::comma(`cases per 100k`, accuracy = 1), " per 100K")), color = "black", size = 3) +
  coord_sf(datum = NA, label_axes = "----") +
  xlab("") + ylab("") + theme_minimal()  +
  # theme(legend.position = "none") +
  labs(title = "Total Cumulative Cases by Connecticut County",
       subtitle = "shading shows total cases per 100K of population as of most recent report\n(without Somers, Brooklyn, and Montville)",
       caption = "Sources: CT DPH, US Census, tidycensus package")

# p_county_map_new_cases <- ggplot() +
#   geom_sf(data = county_geometries %>%
#             left_join(dph_counties %>% filter(date == max(date, na.rm = TRUE)) %>%
#                         select(county, rnew_cases_per1k),
#                       by = "county") %>% 
#             mutate(`new per 100k` = rnew_cases_per1k * 100),
#           # need aes(geometry = geometry) because joined sf to a tibble and so lost sf
#           aes(fill = `new per 100k`, geometry = geometry)) +
#   scale_fill_gradient(low = "white", high = "grey") +
#   geom_sf_text(data = county_centroid, aes(label = paste0(county, "\n", round(rnew_cases_per1k*100, 1), " per 100K")), color = "black", size = 3) +
#   coord_sf(datum = NA, label_axes = "----") +
#   xlab("") + ylab("") + theme_minimal()  +
#   # theme(legend.position = "none") +
#   labs(title = "New Cases (7-Day Rolling Average) by Connecticut County",
#        subtitle = "shading shows new cases per 100K of population as of most recent report",
#        # fill = "cases per 1K",
#        caption = "Sources: CT DPH, US Census, tidycensus package")
# 
# p_county_map_cum_cases <- ggplot() +
#   geom_sf(data = county_geometries %>%
#             left_join(dph_counties %>% filter(date == max(date, na.rm = TRUE)) %>%
#                         select(county, cases, deaths),
#                       by = "county") %>% 
#             mutate(`cases per 100k` = cases / (total_pop / 100000)),
#           # need aes(geometry = geometry) because joined sf to a tibble and so lost sf
#           aes(fill = `cases per 100k`, geometry = geometry)) +
#   scale_fill_gradient(low = "white", high = "grey") +
#   geom_sf_text(data = county_centroid, aes(label = paste0(county, "\n", comma(cases, accuracy = 1.0), " cases\n", comma(deaths, accuracy = 1.0), " deaths")), color = "black", size = 3) +
#   coord_sf(datum = NA, label_axes = "----") +
#   xlab("") + ylab("") + theme_minimal()  +
#   # theme(legend.position = "none") +
#   labs(title = "Cumulative Cases and Deaths by Connecticut County",
#        subtitle = "shading shows total cases per 100K of population",
#        # fill = "cases per 1K",
#        caption = "Source: CT DPH, US Census, tidycensus package")

```
Next I'll look at variations within Connecticut. Below are two maps
showing the eight counties in Connecticut^[The data displayed on these maps excludes the towns of 
Somers, Brooklyn, and Montville because they
are small rural towns with prisons, and the cases reported in 
the prisons may dominate the total in the town and even affect the 
reporting for the county overall. Covid-19 in prisons is a significant
issue, but it's helpful to try to evaluate it separately from 
data from the non-prison community.]. 
The first map in Figure \@ref(fig:county-maps) shows 
the cumulative total of cases since the start of the epidemic.
The second map shows only new cases reporting during the last two weeks.
On both maps, cases are adjusted to show cases per 100K of population.
So the first map relates more to the total impact of the disease over the
full course of the epidemic to date, and the second
map is an indication of the recent prevalence of the
virus. 

Connecticut is part of the epidemic's surge in the New York area and the magnitude
of the epidemic has been greater in the counties closer to New York City. Back in March cases
first appeared in Fairfield County and then spread to Litchfield, New Haven, and Hartford.
More recently Hartford has been the area that we have to watch.

<br/><br/><br/>
```{r county-maps, echo = FALSE, fig.cap = "County Maps of New and of Cumulative Cases", out.width = '50%', fig.show = 'hold'}
print(p_county_map_cum_cases)
print(p_county_map_current_cases)
```
```{r town-categories, echo = FALSE}
by_category <- dph_towns %>% 
  combine_dph_towns(category)
p_by_category <- ggplot(data = by_category,
                        aes(x = date, y = rnew_cases_per1k * 100)) +
  geom_line() +
  facet_wrap(~ category)

```

### By County and Type of Town

There are 169 towns in Connecticut, which makes it hard to examine variations by town. To help
examine variations within counties, I'll use a typology of towns called 
[The Five Connecticuts](https://ctdatahaven.org/sites/ctdatahaven/files/UConnCPR%20Changing%20Demographics-5%20CTs%202004.pdf)
that divides towns into five categories based on census variables: Urban Core, Urban Periphery,
Suburban, Rural, and a fifth category for Wealthy Suburbs used for some towns in Fairfield County. 
Adjusting for population, the number of cases has been greater in the counties closer
to New York City. 
In the figure below I'll fold the wealthy towns into the Suburban category. I have excluded
Montville and Somers because the prisons in those towns complicates interpretation of the
town statistics.

Counties are laid out from top to bottom and the counties closer to New York are toward the top. 
The columns show the four categories. Urban towns have been hit harder than suburban and rural
towns. While density may have some effect on case rates, the report [Towards Health Equity in Connecticut: The Role of Social Inequality and the Impact of COVID-19](https://ctdatahaven.org/reports/towards-health-equity-connecticut)
by DataHaven documents the unequal impact of the epidemic in Connecticut related to social inequality.

```{r county-by-category, echo = FALSE, fig.asp = 1.2}

# create 
for_chart <- dph_towns %>% filter(!(town %in% c("Montville", "Somers", "Brooklyn"))) %>% 
  mutate(category = fct_recode(category, Suburban = "Wealthy")) %>% 
         # category = fct_expand(category, "With Prisons"),
         # category = ifelse(town %in% c("Somers", "Brooklyn", "Montville"), "With Prisons", category)) %>% 
  combine_dph_towns(county, category) %>% 
  mutate(rnew_cases_per100k = rnew_cases_per1k * 100,
    county = factor(county,
      levels = c("Litchfield", "Fairfield", "Hartford", "New Haven", "Middlesex", "Tolland", "New London", "Windham")),
    for_label = paste0(towns, ifelse(towns > 1, " towns", "town"), ", pop=", comma(total_pop, accuracy = 1)))
ggplot(data = for_chart,
                        aes(x = date, y = rnew_cases_per100k)) +
  geom_line() +
  # facet_grid(county ~ category, nrow = 8, drop = FALSE) +
  facet_grid(county ~ category, drop = FALSE) +
  geom_text(data = for_chart %>% filter(date == min(date)),
            x = -Inf, y = Inf,
            aes(label = for_label), hjust = 0, vjust = 1, size = 3) +
  ylab("New Cases per 100K") + xlab(NULL) +
  labs(title = "New Cases by County and by Type of Town Per Population",
       subtitle = "(7-day rolling average of new cases)",
       caption = '"Wealthy" towns are included with "Suburban".\nSomers, Brooklyn, and Montville are omitted because prison data may dominate town data.',
       tag = NULL)

```

### Details by Town

Below are links to a separate web page by for each of the eight counties in Connecticut.
Click on the county name for detail about individual towns within that county. 

   |  |   | 
--------------------- | --------------------- | --------------------- | --------------------- 
[Litchfield County](/post/2020-06-22-tracking-covid-19-in-connecticut/index.en_files/litchfield_summary.html) | [Hartford County](/post/2020-06-22-tracking-covid-19-in-connecticut/index.en_files/hartford_summary.html) | [Tolland County](/post/2020-06-22-tracking-covid-19-in-connecticut/index.en_files/tolland_summary.html) | [Windham County](/post/2020-06-22-tracking-covid-19-in-connecticut/index.en_files/windham_summary.html)
[Fairfield County](/post/2020-06-22-tracking-covid-19-in-connecticut/index.en_files/fairfield_summary.html) | [New Haven County](/post/2020-06-22-tracking-covid-19-in-connecticut/index.en_files/new_haven_summary.html) | [Middlesex County](/post/2020-06-22-tracking-covid-19-in-connecticut/index.en_files/middlesex_summary.html) | [New London County](/post/2020-06-22-tracking-covid-19-in-connecticut/index.en_files/new_london_summary.html) 

************************

## Links to Sources

* State of Connecticut Department of Public Health 
    + Daily Covid-19 Update in the form of a PDF. A [list of the reports](https://data.ct.gov/Health-and-Human-Services/COVID-19-Daily-DPH-Reports-Library/bqve-e8um) is available via Connecticut Open Data.
    + [Connecticut Covid-19 Data Tracker](https://portal.ct.gov/Coronavirus/COVID-19-Data-Tracker)
    + Links to [all the items available via Connecticut Open Data](https://data.ct.gov/stories/s/COVID-19-data/wa3g-tfvc/#data-library) The Connectiut Open Data site has been a pleasure to use. 
    
* [Mapping and Covid-19 at the Yale Medical School](https://covid.yale.edu/innovation/mapping/) I have had an opportunity to
observe some of these efforts for me it has been a great source of information and inspiration.

* [New York Times Tracking the Coronavirus](https://www.nytimes.com/interactive/2020/world/coronavirus-maps.html)

* [rt.live](https://rt.live) I have made heavy use of this site so I'll link to it here as well.

* [Covid-19 Tracking Project](https://covidtracking.com)

## Methodology and Sources 

### Some Notes on the Data Available From the Department of Public Health

Sometimes there is a lag before cases and deaths end up in the daily reports.
As a result, there tends to be a "day of the week" effect.
For that reason, observers of the Covid-19 statistics generally focus
on a 7-day rolling average of the daily counts.
Some of the unusually large peaks and valleys in these charts are due
to reporting process. The DPH has also begun producing reports based
on the [date a sample was taken for a test](https://data.ct.gov/Health-and-Human-Services/Number-of-COVID-19-Cases-by-Date-of-Specimen-Colle/xz44-6swc) 
and [by the date of death](https://data.ct.gov/Health-and-Human-Services/COVID-19-associated-deaths-by-date-of-death/abag-bjkj). 
That's a more accurate way to look at the change over time, but it
means the data for the most recent days are hard to interpret because
some data may be "in process" and not yet reported. In these charts I 
have used "date of report" rather than "date of sample taken" or "date of death"
because that gives me all of the recent data that is available
and because that is what most other data projects (such as the Covid-19 Project
or the New York Times) have been using.

The DPH reports always include a comment that "all data are preliminary and subject
to change." But when they make and adjustment to correct errors, as far as
I can tell they do not go back and adjust the earlier reports. That can
lead to a misleading report of recent changes. For example, in one note
below they removed 70 cases because of errors on a day in which there were
81 new cases. In the data series that I download, that shows up as having
been 11 new cases that day, not 81, because 70 were removed. Earlier days
are not corrected.

[Note as of June 1](https://portal.ct.gov/-/media/Coronavirus/CTDPHCOVID19summary6012020.pdf?la=en)

> *In Connecticut during the early months of this pandemic, it became increasingly clear that it would be necessary to track probable COVID-19 cases and deaths, in addition to laboratory-confirmed (RT- PCR) cases and deaths. This was needed to better measure the burden and impact of this disease in our communities and is now part of the national surveillance case definition for COVID-19. Today for the first time, DPH is reporting cases and deaths as “confirmed” or “probable.” Previous reports
> reported these as a combined number. The only change today is that they are being separated to conform with CDC reporting guidance. Probable cases of COVID-19 involve persons who have not had confirmatory laboratory testing (RT-PCR) performed for COVID-19, but whose symptoms indicate they are very likely to have a COVID-19 infection. In Connecticut, most of the probable COVID-19 cases involve persons whose death certificates list COVID-19 disease or SARS-CoV-2 as a cause of death or a significant condition contributing to death.

[Note as of May 27](https://portal.ct.gov/-/media/Coronavirus/CTDPHCOVID19summary5272020.pdf?la=en)

>The staff at the Department of Public Health have removed 356 cases and 808 tests in the past 24 hours, which were identified as duplicates in the system, affecting both test and overall case numbers. Since yesterday, there have been 341 new positive cases, and 5,215 new tests were reported.


[Note as of June 18](https://portal.ct.gov/-/media/Coronavirus/CTDPHCOVID19summary6182020.pdf)

>Please note that 81 new cases were reported in the past 24-hours; 70 previously reported cases were removed from the total counts due to correction of data errors.

In the data portal, the number of cases reported for Montville was 381 on June 17.
As of June 18 it reports 293 cases, a reduction of 88 cases. That's more than
the 70 cases removed. Perhaps the case counts for Montville were affected by
inmates being moved within the Connecticut prison system.

[Note as of June 24, 2020](https://portal.ct.gov/-/media/Coronavirus/CTDPHCOVID19summary6242020.pdf)

>1175 new test results were reported since the last report and 2770 previously reported PCR tests were removed due to correction of data errors.